Finding Maximum $k$-Cliques Faster Using Lazy Global Domination

### 2016-05-17 paper/paper.tex
Address review 2

A clique in a graph is a set of vertices, each of which is adjacent to every other vertex in the
set. Finding a clique of maximum size in a graph is one of the basic \NP-hard problems
\cite{Garey:1990}; applications include geometry, coding theory, computer vision and bioinformatics
\cite{Bomze:1999,Butenko:2006}. However, when analysing real-world data, a clique may be too strong
a requirement. A $k$-clique (or sometimes $n$-clique or $s$-clique---in an unfortunate clash of
notation, ``$k$-clique problem'' is sometimes used elsewhere for the decision version of the clique
problem, to distinguish it from the maximum clique problem) is a relaxed form of clique, where
instead of requiring each pair of vertices to be directly adjacent, we only require they be
connected by a path of length at most $k$ \cite{Luce:1950}.  Thus a $1$-clique \emph{is} a clique, a
$2$-clique may be thought of as ``a group of people, all of whom either know each other or have a
mutual acquaintance'', and so on. We illustrate this in \cref{figure:k-cliques}.  Determining the
size of a maximum $k$-clique is \NP-hard for any fixed $k$ \cite{Bourjolly:2002}.

@@

A maximum clique algorithm can easily be adapted to find a maximum $k$-clique in a graph $G$ by
considering the graph $G^k$, which we describe below. However, it is not obvious that this is a
viable approach: even if $G$ is sparse, $G^k$ may not be, and the maximum clique problem on dense
graphs can be very challenging computationally. Here we take a state-of-the-art maximum clique
algorithm which is suitable for use on dense graphs \cite{Prosser:2012}, and
investigate whether this approach is feasible in practice. We modify the algorithm to include a new
lazy ``global domination'' inference step---this technique provides no benefit for typical maximum
clique problems, but for maximum $k$-clique graphs it sometimes gives improvements of several orders
of magnitude. We present computational results for the maximum $k$-clique problem on a range of
benchmark and real-world graphs. We finish with a detailed look at random graphs.

@@

Our approach for finding a maximum $k$-clique is presented as \cref{algorithm:maxKClique}. Our first
step (\mcline{power}) is to replace our input graph $G$ with a modified graph $G^k$. This graph has
the same vertex set as $G$, and edges between any two distinct vertices $v_1$ and $v_2$ iff there is
a path of length at most $k$ between $v_1$ and $v_2$ in $G$. We may construct this graph using a
bounded breadth-first search: we refer to \citeauthor{Chang:2013} \shortcite{Chang:2013} for how to
implement this quickly in practice. Now it is easy to see that maximum cliques in $G^k$ correspond
with maximum $k$-cliques in $G$ \cite{Balasundaram:2005}.

@@

By comparing these results with the $k$-club results of Chang et al.\ \shortcite{Chang:2013}, we see that
in all but four cases the $k$-clique and $k$-club numbers are equal; all of these differences occur
when $k = 4$. (Chang et al.\ did not investigate the ``Erdos02'' graph, but Wotzlaw
\shortcite{Wotzlaw:2014} confirmed privately that the $k$-clique and $k$-club numbers are the same here
too.) On the other hand, the $k$-clique numbers are sometimes much easier to find, both
algorithmically and computationally.

>>>>

A clique in a graph is a set of vertices, each of which is adjacent to every other vertex in the
set. Finding a clique of maximum size in a graph is one of the basic \NP-hard problems
\cite{Garey:1990}; applications include geometry, coding theory, computer vision and bioinformatics
\cite{Bomze:1999,Butenko:2006}. However, when analysing real-world data, a clique may be too strong
a requirement. A $k$-clique (or sometimes $n$-clique or $s$-clique---in an unfortunate clash of
notation, ``$k$-clique problem'' is sometimes used elsewhere for the decision version of the clique
problem, to distinguish it from the maximum clique problem) is a relaxed form of clique, where
instead of requiring each pair of vertices to be directly adjacent, we only require that they be
connected by a path of length at most $k$ \cite{Luce:1950}.  Thus a $1$-clique \emph{is} a clique, a
$2$-clique may be thought of as ``a group of people, all of whom either know each other or have a
mutual acquaintance'', and so on. We illustrate this in \cref{figure:k-cliques}.  Determining the
size of a maximum $k$-clique is \NP-hard for any fixed $k$ \cite{Bourjolly:2002}.

@@

A maximum clique algorithm can easily be adapted to find a maximum $k$-clique in a graph $G$ by
considering the graph $G^k$: this graph has the same vertex set as $G$, and edges between any two
distinct vertices $v_1$ and $v_2$ iff there is a path of length at most $k$ between $v_1$ and $v_2$
in $G$. It is easy to see that maximum cliques in $G^k$ correspond with maximum $k$-cliques in $G$
\cite{Balasundaram:2005}.  However, it is not obvious that this is a viable approach: even if $G$ is
sparse, $G^k$ may not be, and the maximum clique problem on dense graphs can be very challenging
computationally. Here we take a state-of-the-art maximum clique algorithm which is suitable for use
on dense graphs \cite{Prosser:2012}, and investigate whether this approach is feasible in practice.
We modify the algorithm to include a new lazy ``global domination'' inference step---this technique
provides no benefit for typical maximum clique problems, but for maximum $k$-clique graphs it
sometimes gives improvements of several orders of magnitude. We present computational results for
the maximum $k$-clique problem on a range of benchmark and real-world graphs. We finish with a
detailed look at random graphs.

@@

Our approach for finding a maximum $k$-clique is presented as \cref{algorithm:maxKClique}. Our first
step (\mcline{power}) is to replace our input graph $G$ with $G^k$. We may construct this graph
using a bounded breadth-first search: we refer to \citeauthor{Chang:2013} \shortcite{Chang:2013} for
how to implement this quickly in practice.

@@

By comparing these results with the $k$-club results of Chang et al.\ \shortcite{Chang:2013}, we see
that in all but four cases the $k$-clique and $k$-club numbers are equal; all of these differences
occur when $k = 4$. (Chang et al.\ did not investigate the ``Erdos02'' graph, but
\citeauthor{Wotzlaw:2014} confirmed privately that the $k$-clique and $k$-club numbers are the same
here too.) On the other hand, the $k$-clique numbers are sometimes much easier to find, both
algorithmically and computationally.
